---
title: "R Programming: An Introduction"
author: "Stacey Borrego (edited by Kristina Riemer)"
output:
  pdf_document: 
      highlight: zenburn
  word_document: default
  html_document: default
fontsize: 12pt
urlcolor: blue
linkcolor: red
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Helpful Links

-   Data Carpentry: [Data Analysis and Visualization in R for
    Ecologists](https://datacarpentry.org/R-ecology-lesson/index.html)
-   [Tidyverse](https://tidyverse.org/): R packages for Data Science
-   [R for Data Science 2nd Edition](https://r4ds.hadley.nz/) by Hadley
    Wickham and Garret Grolemund
-   [Advanced R](http://adv-r.had.co.nz/) by Hadley Wickham
-   [ggplot2: Elegant graphics for data
    analysis](https://ggplot2-book.org/) by Hadley Wickham
-   [Posit Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
    by Posit
-   [The R
    Gallery](https://bookdown.org/content/b298e479-b1ab-49fa-b83d-a57c2b034d49/)
    by Kyle W. Brown
-   [Introduction to R](https://intro2r.com/) by Douglas, Roos, Mancini,
    Cuoto, & Lusseau

## Notes

- Sections marked with ^\$^ may have been skipped due to time limits

## Overview of RStudio

-   (Lower) left hand side: *console* which runs R code
-   Create a new script
    -   Hit file button in upper left hand corner
    -   Select first option "R Script"
-   Upper left hand side: *scripts* where R code can be saved
-   Upper right hand side: where objects are
-   Lower right hand side: files and folders on computer

## Simple math

In the *console*, type in the following expressions\
Hit `ENTER` after each expression

```{r eval=FALSE}
3 + 5
12 / 7
3 * 4
8 ^ 2
3 + 5 / 2
(3 + 5) / 2
```

## Creating objects in R

### Working with variables/objects

Do this in the *script* that was created

Assign values to a variable/object

```{r}
weight_kg <- 55
```

### Running code in a script

-   Select `Run` button at the top of the script window
-   Select code or move cursor to the line to be run and type `Ctrl` +
    `ENTER`

**How to name variables:**

-   Assignment operator (\<-)
-   Cannot start with numbers
-   Case sensitive
-   Avoid function names
-   Avoid dots (.)
-   Be consistent in the styling of your code
    -   [R Style Guide](https://google.github.io/styleguide/Rguide.html)
    -   Styles can include “lower_snake”, “UPPER_SNAKE”,
        “lowerCamelCase”, “UpperCamelCase”, etc
    -   We will use "lower_snake" for readibility during this workshop

Print the value of an object

```{r eval=FALSE}
weight_kg
```

Simple math with objects

```{r eval=FALSE}
2.2 * weight_kg
```

Reassign object's value

```{r eval=FALSE}
weight_kg

weight_kg <- 65
weight_kg

2.2 * weight_kg
```

Assign a new object a value containing another object

```{r eval=FALSE}
weight_lb <- 2.2 * weight_kg
weight_lb
```

#### [Exercises]{.underline}

1.  Create a variable for the length of a line and store the value 15.3
2.  Multiply that object by 3 to get volume
3.  Create a variable that stores that volume value

```{r results='hide'}
length <- 15.3
length * 3
volume <- length * 3
```

What is the value of volume?

```{r}
print(volume)
```

### Saving code in a script

-   Save the file
    -   File \> Save (`Ctrl` + `S`)



### Add comments in a script

-   The comment character in R is `#`. Anything to the right of a `#` in
    a script will be ignored.
-   Comment out a section by selecting and pressing `Ctrl` + `Shift` +
    `C`
    -   press the same keys again to remove the comment

```{r eval=FALSE}
# This is a comment.
# The sum of 2 + 2
2 + 2
```

### Functions and arguments

-   Functions are scripts that automate more commands
-   Functions are available in base R, by importing packages, or by
    making them yourself
-   Usually require one or more inputs called **arguments**
-   Functions typically return a **value**

```{r eval=FALSE}
# This is a function call
sqrt(10)
weight_kg <- sqrt(10)
weight_kg
```

-   Information about functions can be found in the `Help` window by
    typing `?` and the function name. Example: `?round`
-   You can also use the function `args()` to see the arguments of a
    function.

```{r eval=FALSE}
round(3.14159)

?round
args(round)

round(3.14159, digits = 2)
```

#### [Exercises]{.underline}

1.  Get the absolute value of -253 (hint: look at at the `Help` page for
    the `sqrt` function)
2.  Take the logarithm of 10 (hint: the function name is the first three
    letters of the word)
3.  Take the logarithm of 10, but change the base to `exp(3)`

```{r}
abs(-253)
log(10)
log(10, base = exp(3))
```

## Vectors and data types

-   A vector is composed by a series of values, which can be either
    numbers or characters.
    -   Quotes must surround characters.
    -   In a vector, all of the elements are the same type of data.
-   We can assign a series of values to a vector using the `c()`
    function.

```{r eval=FALSE}
# Numerical vector
weight_g <- c(50, 60, 65, 82)
weight_g

# Character vector
animals <- c("mouse", "rat", "dog")
animals
```



### Explore the content of vector

-   `length()` tells you how many elements are in a particular vector
-   `class()` indicates what kind of object you are working with
-   `str()` provides an overview of the structure of an object and its
    elements.

```{r eval=FALSE}
# Length
length(weight_g)
length(animals)

# Class
class(weight_g)
class(animals)

# Structure
str(weight_g)
str(animals)
```

#### [Exercises]{.underline}

1.  Create a character vector of names of colors
2.  Put the vector elements into alphabetical order using the `sort`
    function

```{r}
colors <- c("red", "blue", "purple", "black")
sort(colors)
```

### ^\$^Add elements to a vector

```{r eval=FALSE}
# Add to the end of the vector
weight_g <- c(weight_g, 90)

# Add to the beginning of the vector
weight_g <- c(30, weight_g) 

# Inspect the modified vector
weight_g
```

#### [Exercises]{.underline}

1.  Add another color to your colors vector
2.  Put the vector into reverse alphabetical order (hint: use `?`)

```{r}
colors <- c(colors, "magenta")
sort(colors, decreasing = TRUE)
```

### Data Types

An atomic vector is the simplest R data type and is a linear vector of a
single type.

There are four common types of atomic vectors: character, numeric or
double, logical, and integer. There are two rare types that will not be
discussed: complex and raw.

-   `character`
-   `numeric` or `double`
-   `logical` for TRUE and FALSE (the boolean data type)
-   `integer` for integer numbers (e.g., 2L, the L indicates to R that
    it’s an integer)\
-   `complex` to represent complex numbers with real and imaginary parts
    (e.g., 1 + 4i)
-   `raw` for bitstreams

```{r eval=FALSE}
typeof(weight_g)
typeof(animals)
```

### Coercion

All elements of an atomic vector must be the same type, so when you
attempt to combine different types they will be coerced to the most
flexible type. Types from least to most flexible are: logical, integer,
double, and character.

```{r eval=FALSE}
num_char <- c(1, 2, 3, "a")
typeof(num_char)
str(num_char)

num_logical <- c(1, 2, 3, TRUE, FALSE)
typeof(num_logical)
str(num_logical)

char_logical <- c("a", "b", "c", TRUE)
typeof(char_logical)
str(char_logical)

tricky <- c(1, 2, 3, "4")
typeof(tricky)
str(tricky)
```

#### [Exercises]{.underline}

1.  Create a logical vector with any number and order of `TRUE` and
    `FALSE`
2.  Add a numeric value to this vector
3.  What kind of vector will this be?

```{r}
some_booleans <- c(TRUE, FALSE, FALSE, FALSE)
some_booleans <- c(3, some_booleans)
typeof(some_booleans)
```

## ^\$^Conditional subsetting

To extract one or several values from a vector, provide one or several
indices in square brackets.

```{r eval=FALSE}
animals <- c("mouse", "rat", "dog", "cat")

animals[2]

animals[c(3, 1)]

animals[1:3]

more_animals <- animals[c(1, 4, 1, 3, 1, 2)]
more_animals

```

Subset by using a logical vector. TRUE will select the element with the
same index, while FALSE will not.

```{r eval=FALSE}

weight_g <- c(21, 34, 55)

weight_g[c(TRUE, FALSE, TRUE)]

weight_g > 50

weight_g[weight_g > 50]
```

Combine multiple tests using `&` (both conditions are true, **AND**) or
`|` (at least one of the conditions is true, **OR**).

Helpful operators

-   `>` greater than
-   `<` less than
-   `<=` less than or equal to
-   `==` equal to, test for numerical equality between the left and
    right hand sides

```{r eval=FALSE}
weight_g

weight_g[weight_g > 20 & weight_g < 30]

weight_g[weight_g <= 30 | weight_g == 55]

weight_g[weight_g >= 30 & weight_g == 21]
```

Search for strings in a vector

```{r eval=FALSE}
animals <- c("dog", "rat", "cat", "cat")
animals

animals[animals == "cat" | animals == "rat"]

animals %in% c("rat", "cat", "duck", "fish")

animals[animals %in% c("rat", "cat", "duck", "fish")]

```



## Missing data

-   Missing data are represented in vectors as NA.
-   You can add the argument `na.rm = TRUE` to calculate the result as
    if the missing values were removed (rm stands for **r**e**m**ove)
-   `is.na()` identifies missing elements
-   `na.omit()` returns the object with incomplete cases removed
-   `complete.cases()` returns a logical vector indicating which cases
    are complete

```{r eval=FALSE}
heights <- c(1, 2, 3, NA, 5, NA)

# Performing an operation on a vector with NA will usually return NA
mean(heights)
max(heights)

# Using `na.rm = TRUE` will exclude NA values and perform the operation
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

### ^\$^Function examples

```{r eval=FALSE}
# Returns vector of boolean values indicating if the value NA or not
is.na(heights)

# Sum of all the NAs present
sum(is.na(heights))

# Identifies the index containing the NA value
which(is.na(heights))

# Returns the TRUE values for `is.na()`
heights[is.na(heights)]

# Returns the FALSE values for `is.na()`
heights[!is.na(heights)]

# Returns values with incomplete cases (NA) removed
na.omit(heights)
mean(na.omit(heights))

# Returns values with complete cases only - NA removed
complete.cases(heights)
heights[complete.cases(heights)]
```



## Data Structures

Vectors are one of the many data structures that R uses.

-   lists (`list`)
-   matrices (`matrix`)
-   data frames (`data.frame`)
-   factors (`factor`)
-   arrays (`array`)

### Data Frames

When we load data into R, it may be stored as an object of class data
frame.

A data frame is the representation of data in the format of a table
where the columns are vectors that all have the same length. Because
columns are vectors, each column must contain a single type of data
(e.g., characters, integers, factors, logical values).

For example, here is an example data frame comprising a numeric,
character, and logical vector.

```{r eval=TRUE, echo=TRUE}
df <- data.frame(values = c(1, 3, 5, 7, 9, 11),
                 some_characters = c("F", "O", "O", "B", "A", "R"),
                 true_false = c(TRUE, FALSE, TRUE, FALSE, FALSE, FALSE))
df
```

### Simple column subsetting

Data frames can also be subset using square brackets as demonstrated
with the single vectors or using `$` to specifically name a column.

Using single square brackets will return a data frame containing the
values of the vector, whereas using double square brackets or `$` will
return a simplified vector of the type of class as the values contained
with in it.

```{r}
# Column 1 data frame
df[1]

# Column 1 simplified vector
df[[1]]
df$values
```

Data can be further subset by providing the indices for the rows and
columns you want. The notation is to first name the data frame followed
by single square brackets listing the rows and columns.

```{r}
# Column 2
df[2]
```

### ^\$^Row and column subsetting

```{r}
# Rows 4 through 6 in column 2
df[4:6, 2]

# Rows 1, 5, and 4 in column 2
df[c(1, 5, 4), 2]

# Rows 1, 5, and 4 in all columns
df[c(1, 5, 4), ]

# Rows 1, 5, and 4 in columns 1 through 2
df[c(1, 5, 4), 1:2]

# Rows 1, 5, and 4 in columns 1 and 3
df[c(1, 5, 4), c(1:3)]
```



### ^\$^Column name subsetting

The subset data frame can be further subset using single square brackets
and the specific names of the columns.
[HERE](https://r-coder.com/subset-r/) is a pretty thorough article on
subsetting.

```{r eval=TRUE}
df["some_characters"]
df[1, "some_characters"]
df[1:3, "some_characters"]
df[c(1, 5, 6), c("values", "some_characters")]
```



### Explore the content of dataframe

- `dim()` returns number of rows and columns
- `colnames()` returns names of all columns
-   `str()` can be used with dataframe as with vectors to show structure
- `summary()` returns summary statistics of columns

```{r}
dim(df)
colnames(df)
str(df)
summary(df)
```

